A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-04-13 Manuscript/ohyou.tex
More eidts and moved out Bayesian stuff

  The popularity of the OU model has grown extensively in recent years (Fig. \ref{figure.literature}); over 2500 ecology, evolution and palaeontology papers containing the phrase ``Ornstein Uhlenbeck'' were published between 2012 and 2014 (Google Scholar search 15th March 2015; see Supplementary Material). % Need to add ref to supp mat
  This may partly be because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}. 
  Additionally, although the OU model is pattern-based, it has a number of attractive biological interpretations. 
  For example, fit to an OU model is used as evidence for the action of processes such as phylogenetic niche conservatism, convergent evolution and stabilising selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}. 
  Unfortunately, although widely used, the properties of the OU model are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

@@

  Uses of the OU model can be divided into three broad categories. % improve citations in this section? % fix up numbers. Argggggggghhhhhhhhhhh.
    \begin{enumerate}
      \item \textbf{Modeling the evolution of a continuous character (N = 58).}\\ 
            Generally in these analyses, multiple models of evolution are fit to the same continuous character \citep[e.g.,][]{Harmon:2010aa}. Model selection is then used to determine which model best fits the data. 
            If the OU model is favoured, this is often used as evidence of evolutionary constraints, stabilising selection, or phylogenetic niche conservatism \citep[e.g.,][]{Harmon:2010aa,cooper2010body,Wiens:2010aa,christin2013anatomical}.
      \item \textbf{Modeling the effects of different hypothesised adaptive regimes on a continuous character (N = 27).}\\
            OU models can be fit with one optimal trait value, or multiple different optima \citep{Butler:2004aa}. 
            The latter represents evolution under multiple selective regimes, and is probably more biologically realistic. 
            In these analyses, OU models with different numbers of optima are fitted to the same continuous character, then model selection is used to determine what number of selection regimes best fits the evolution of the trait \citep[e.g.,][]{beaulieu2012modeling}.
            Note that the first and second uses of the OU model are often combined, so that OU models with various numbers of optima are included in the pool of models being tested (N = 27).
      \item \textbf{Controlling for phylogeny in correlative analyses (N = 35).}\\ 
            Phylogenetic generalised least squares (PGLS) models incorporate information about the relationships among species into the error term of a generalised least squares model. 
            This error term generally consists of a variance-covariance matrix of the phylogeny, but various transformations are used (e.g., Pagel's $\lambda$; \citealp{Pagel:1997aa}) to improve the fit of the model to the data. 
            In these analyses, the OU model is used to transform the tree and the variance-covariance matrix.
            This is rarely interpreted as corresponding to any kind of process, it just improves the fit of the PGLS models \citep[e.g.,][]{blankers2012ecological}.
    \end{enumerate}

  \noindent
  The OU model is also used to reconstruct ancestral states \citep{martins1999estimation}
  and to detect clade-wide convergent evolution \citep{ingram2013surface}. 
  The majority of papers use the OU model to model the evolution of a continuous character (Fig 2 and Supplementary Material), % Need to add ref to supp mat
  therefore, we focus on this use of the OU model in our simulations below. 
  The principles here also apply to a range of other macroevolutionary models that can be fit to continuous data and compared using model testing procedures.

    \begin{table}[h]
      \begin{centering}
        \caption{The most common uses of Ornstein Uhlenbeck models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2013. See Supplementary Material for details linkto supp mat.}
        \bigskip  
          \begin{tabular}{p{8cm}lc}
            \hline
            Use of OU model & Optima & Number of papers\\
            \hline
            Ancestral state reconstruction & single & 8\\
            & multiple & 2\\
            Convergent evolution & single & 0\\
            & multiple & 2\\
            Model of evolution & single & 31\\
            & multiple & 27\\
            Phylogenetic generalised least squares & single & 35\\
            & multiple & 0\\
            Other & single & 5\\
            & multiple & 5\\
            Total & single &  36\\
            & multiple & 79\\
            \hline
            \label{table.uses}
          \end{tabular}
    \end{table}

@@

    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large, all imprint of history is lost and the trait evolution is essentially a rapid burst at the present (FIGURE 2). % Fix the figure or send to supplemental

    %$\alpha$ can most easily be interpreted by using it to estimate the ``phylogenetic half-life'' ($t_\frac{1}{2}$) of a trait, i.e., the time it takes for a species entering a new niche to evolve halfway toward its new expected optimum \citep{hansen1997stabilizing}, as follows:

@@

    %If $t_\frac{1}{2}$ is short relative to the branch lengths of the phylogeny, adaptation is quick, residual phylogenetic correlations are weak, and there is little influence of the past. This is sometimes referred to as ``phylogenetic inertia'' but we dislike this term

@@

     % This maybe belongs in the main text in shorter form  
  %Figure \ref{figure.traitsim} gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure \ref{figure.traitsim} highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. \ref{figure.traitsim}B).  Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. \ref{figure.traitsim}C). This will be indistinguishable from phylogenetically unstructured data.

\begin{figure}
\centering
    \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OU_figure2.pdf}
\caption{Scaling of expected trait similarity with time since evolutionary divergence predicted by the Ornstein-Uhlenbeck model. The covariance between species’ traits values is scaled by the intra-specific trait variance (i.e. equal to correlation between species’ traits). This is plotted against the relative time of shared history (time at which species branched from each other, divided by the total tree height). }
\label{figure.traitsim}
\end{figure}

%In terms of statistical model fitting, a feature to highlight is that in Figure \ref{figure.traitsim}A and Figure \ref{figure.traitsim}C, changing the value of α across large relative ranges at the extremes (0 – 0.5 in Fig. \ref{figure.traitsim}A; 10-50 in Fig. \ref{figure.traitsim}C) produces relatively little quantitative change in the model behavior, and it is in the middle region of parameter space (1 – 5) that the biggest discernible effects occur.

%The point of highlighting these aspects of expected behavior is that we believe these properties need to be understood if we are to build up a full picture of when the model is likely to be well estimated, as well as to understand the bounds of what can be inferred from real data.

  \subsection{Maximum Likelihood simulations}
  \label{section:sims.methods} 
    We used simulations to explore the statistical properties of $\alpha$. 
    We simulated phylogenies with 25, 50, 100, 150, 200, 500, or 1000 tips under pure birth, constant-rate Birth-Death (extinction fractions of 0.25, 0.5 and 0.75), or temporally varying speciation rate (speciation rate modeled as time from the root raised to the power 0.2, 0.5, 2 and 5) models. 
    We simulated 1000 phylogenies for each combination of tips and models resulting in 56,000 simulated phylogenies in total. 
    Trees were simulated using the R package TESS \cite{hohna2013fast}. 
    We then simulated the evolution of a single trait under a Brownian motion model on each phylogeny using the R package MOTMOT \citep{Thomas:2011aa}. 
    We estimated $\alpha$ and compared the fit of a Brownian model to that of an OU model using a likelihood ratio test with 1 degree of freedom with the \texttt{transformPhylo.ML} function in MOTMOT \citep{Thomas:2011aa}. 
    This procedure ... % explain WHY we are doing this, how does it relate to uses of the OU model.

@@

  \subsection{Bayesian simulations} % keep? or move it all to supp mat?
    For completeness, we repeated the Maximum Likelihood simulations in a Bayesian framework implemented in BayesTraits. 
    The choice of priors for OU models has not been fully explored. 
    Therefore, we used three alternative sets of priors on $\alpha$: (i) an exponential distribution with mean = 1; (ii), an exponential distribution with mean = 10, and (iii) a uniform distribution bounded at 0 and 20. 
    For all analyses we used a uniform -100 to 100 prior for $\mu$ and uniform 0 to 100 prior for $\sigma^2$. 
    We present the results from the exponential prior with mean = 10 in the main text since this is a broad, liberal prior, but we provide results derived from the other priors in the Supplementary Material (Figs. S2-S5). % add links to this
    We ran the MCMC chains for $1x10^6$ iterations, disregarding the first $1x10^4$ as burn-in. 
    Following burn-in the chains were sampled every 1000 to ensure independence of each consecutive sample. 
    Multiple independent chains were run for each analysis to ensure convergence was reached. 
    We determined fit to an OU model using Bayes factors estimated from a stepping stone sampling procedure \citep{xie2010improving}implemented in BayesTraits. 
    The marginal likelihoods of the models were calculated using a stepping stone sampler in which fifty stones were drawn from a beta distribution with (alpha = 0.4 and beta = 1). 
    Each stone was sampled for 20000 iterations (with the first 5000 iterations discarded). 
    We treated Bayes factors $>$ 2 as evidence favouring the OU model.

@@

\section{Results}
  \subsection{Simulations}
  \label{section:sims.results}

>>>>

  The popularity of the OU model has grown extensively in recent years (Fig. \ref{figure.literature}); over 2500 ecology, evolution and palaeontology papers containing the phrase ``Ornstein Uhlenbeck'' were published between 2012 and 2014 (Google Scholar search 15th March 2015; see Supplementary Material). % Need to add ref to supp mat
  This may partly be because these models are now easy to implement apply via packages in R \citep[e.g. ouch, GEIGER and OUwie;][]{Butler:2004aa,Harmon:2008aa,beaulieu2012ouwie}. 
  Additionally, although the OU model is pattern-based, it has a number of attractive biological interpretations. 
  For example, fit to an OU model is used as evidence for processes such as phylogenetic niche conservatism, convergent evolution and stabilising selection \citep[e.g.,][]{Wiens:2010aa,christin2013anatomical,ingram2013surface}. 
  Unfortunately, although widely used, the properties of the OU model are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

@@

  The OU model is most commonly used to model the evolution of a continuous character (Table \ref{table.uses}; Supplementary Material). 
  Multiple models of evolution \citep[e.g., Brownian motion, OU, Early burst etc.;][]{Harmon:2010aa,cooper2010body} are fit to the same continuous character and model selection is then used to determine which model best fits the data.  
  OU models can be fit with one optimal trait value, or multiple different optima \citep{Butler:2004aa,beaulieu2012modeling}. 
  The latter represents evolution under multiple selective regimes, and may be more biologically realistic. 
  OU models with various numbers of optima are often included in the pool of evolutionary models being compared \citep[e.g.,][]{} (Table \ref{table.uses})

  OU models are also commonly used to control for phylogeny in correlative analyses (Tables \ref{table.uses}; Supplementary Material).  
  Phylogenetic generalised least squares (PGLS) models incorporate information about the relationships among species into the error term of a generalised least squares model. 
  This error term generally consists of a variance-covariance matrix of the phylogeny, but various transformations are used (e.g., Pagel's $\lambda$; \citealp{Pagel:1997aa}) to improve the fit of the model to the data. 
  The $\alpha$ parameter from an OU model can also be used to transform the tree and the variance-covariance matrix.
  This is rarely interpreted as corresponding to any kind of process, it just improves the fit of the PGLS models \citep[e.g.,][]{blankers2012ecological}. Finally, the OU model is also used to reconstruct ancestral states \citep{martins1999estimation} and to detect clade-wide convergent evolution \citep{ingram2013surface}.

  The majority of papers use the OU model to model the evolution of a continuous character (Table \ref{table.uses}; Supplementary Material) therefore, we focus on this use of the OU model in our simulations below. 
  The principles here also apply to a range of other macroevolutionary models that can be fit to continuous data and compared using model testing procedures.
  Note that we focus on OU models with a single optimum trait value because these are more commonly used (Table \ref{table.uses}; Supplementary Material) and easier to simulate, but our general conclusions also apply to multiple optima OU models.

    \begin{table}[h]
        \caption{The most common uses of Ornstein Uhlenbeck models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2013. See Supplementary Material for details.}
        \bigskip  
          \begin{tabular}{p{8cm}lc}
            \hline
            Use of OU model & Optima & Number of papers\\
            \hline
            Ancestral state reconstruction & single & 8\\
            & multiple & 2\\
            Convergent evolution & single & 0\\
            & multiple & 2\\
            Model of evolution & single & 31\\
            & multiple & 27\\
            Phylogenetic generalised least squares & single & 35\\
            & multiple & 0\\
            Other & single & 5\\
            & multiple & 5\\
            Total & single &  36\\
            & multiple & 79\\
            \hline
            \label{table.uses}
          \end{tabular}
    \end{table}

@@

    When $\alpha$ is close to zero, evolution is approximately Brownian, then as $\alpha$ gets larger the non-Brownian behaviour of the model starts to become apparent. 
    Eventually, when $\alpha$ is really large, all imprint of history is lost and the trait evolution is essentially a rapid burst at the present.

    $\alpha$ can most easily be interpreted by using it to estimate the ``phylogenetic half-life'' ($t_\frac{1}{2}$) of a trait, i.e., the time it takes for a species entering a new niche to evolve halfway toward its new expected optimum \citep{hansen1997stabilizing}, as follows:

@@

    This is sometimes referred to as ``phylogenetic inertia'' but we dislike this term because it means different things to different researchers \citep{Losos:2008aa}.
    If $t_\frac{1}{2}$ is short relative to the branch lengths of the phylogeny, evolution towards the optimum trait value is fast, residual phylogenetic correlations are weak, and there is little influence of the past on trait values \citep{hansen1997stabilizing}.

@@

    \begin{enumerate}
      \item $\alpha$ is bounded at zero.
      \item blah
      \item measurement errors
    \end{enumerate}

  \subsection{Maximum Likelihood simulations}
  \label{section:sims.methods} 
    We used simulations to explore the statistical properties of $\alpha$. 
    We simulated phylogenies with 25, 50, 100, 150, 200, 500, or 1000 tips under pure birth, constant-rate Birth-Death (extinction fractions of 0.25, 0.5 and 0.75), or temporally varying speciation rate (speciation rate modeled as time from the root raised to the power 0.2, 0.5, 2 and 5) models. 
    We simulated 1000 phylogenies for each combination of tips and models resulting in 56,000 simulated phylogenies in total. 
    Trees were simulated using the R package TESS \cite{hohna2013fast}. 
    We then simulated the evolution of a single trait under a Brownian motion model on each phylogeny using the R package MOTMOT \citep{Thomas:2011aa}. 
    We estimated $\alpha$ and compared the fit of a Brownian model to that of an OU model using a likelihood ratio test with 1 degree of freedom with the \texttt{transformPhylo.ML} function in MOTMOT \citep{Thomas:2011aa}. 
    This procedure mirrors a situation where the OU model  %%% argh

@@

@@

\section{Results}
  \subsection{Maximum Likelihood simulations}
  \label{section:sims.results}

### 2015-04-13 Manuscript/ohyou_suppmat.tex
Edits to intro and fixed equations
More eidts and moved out Bayesian stuff

    Next we filtered our Google Scholar search results to focus on empirical papers using OU models (rather than methods papers with empirical proof of concept) published in the following journals: The American Naturalist, Ecology Letters, Evolution, Journal of Evolutionary Biology, Nature, Proceedings of the National Academy of Sciences, USA, Proceedings of the Royal Society B: Biological Sciences, and Science. 
    For each of these papers we recorded the number of species in the analysis, the study group (amphibians, birds, fish, mammals, reptiles, invertebrates or plants), the statistical package or specific R package used to fit the models, and the reason the authors state for using an OU model (ancestral state reconstructions, detecting convergent evolution, controlling for phylogeny, selecting a model of trait evolution, or other). 
    Where papers included multiple analyses using different numbers of species we used the median number of taxa. 
    Where papers had multiple study groups, statistics/R packages or reasons for fitting OU models we counted them in each relevant category. 
    We summarise these results in Fig. %figure link
    and the full dataset is available in Supplemental Table S1 along with the full list of references.

    In total, our literature search found 3720 papers published between 2005 and 2014, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). 
    Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). 
    The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

>>>>

    Next we filtered our Google Scholar search results to focus on empirical papers using OU models (rather than pure methods papers) published in the following journals: The American Naturalist, Ecology Letters, Evolution, Journal of Evolutionary Biology, Nature, Proceedings of the National Academy of Sciences, USA, Proceedings of the Royal Society B: Biological Sciences, and Science.
    We only include papers up to the end of 2013 to ensure completeness.

    For each of these papers we recorded the number of species in the analysis, the study group (amphibians, birds, fish, mammals, reptiles, invertebrates or plants), the statistical package or specific R package used to fit the models, and the reason the authors state for using an OU model (ancestral state reconstructions, detecting convergent evolution, controlling for phylogeny, selecting a model of trait evolution, or other). 
    Where papers included multiple analyses using different numbers of species we used the median number of taxa. 
    Where papers had multiple study groups, statistics/R packages or reasons for fitting OU models we counted them in each relevant category. 
    We summarise these results in Figure 1 and Table 1, and the full dataset is available in Supplemental Table S1 % create
    along with the full list of references.

    In total, our literature search found 3720 papers published between 2005 and 2014, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). 
    Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). 
    The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the ``best'' model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

  \subsection{Bayesian simulations} % keep? or move it all to supp mat?
    For completeness, we repeated the Maximum Likelihood simulations in a Bayesian framework implemented in BayesTraits. 
    The choice of priors for OU models has not been fully explored. 
    Therefore, we used three alternative sets of priors on $\alpha$: (i) an exponential distribution with mean = 1; (ii), an exponential distribution with mean = 10, and (iii) a uniform distribution bounded at 0 and 20. 
    For all analyses we used a uniform -100 to 100 prior for $\mu$ and uniform 0 to 100 prior for $\sigma^2$. 
    We present the results from the exponential prior with mean = 10 in the main text since this is a broad, liberal prior, but we provide results derived from the other priors in the Supplementary Material (Figs. S2-S5). % add links to this
    We ran the MCMC chains for $1x10^6$ iterations, disregarding the first $1x10^4$ as burn-in. 
    Following burn-in the chains were sampled every 1000 to ensure independence of each consecutive sample. 
    Multiple independent chains were run for each analysis to ensure convergence was reached. 
    We determined fit to an OU model using Bayes factors estimated from a stepping stone sampling procedure \citep{xie2010improving}implemented in BayesTraits. 
    The marginal likelihoods of the models were calculated using a stepping stone sampler in which fifty stones were drawn from a beta distribution with (alpha = 0.4 and beta = 1). 
    Each stone was sampled for 20000 iterations (with the first 5000 iterations discarded) 
    We treated Bayes factors $>$ 2 as evidence favouring the OU model.
